Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-08 aistats.tex
Cast MAP as SMD

\newcommand{\MAPm}{\hat \m_n}
\newcommand{\MAPt}{\hat \nat_n}

@@

There are algorithms for this setting, for example the MLE -- the (online) maximum likelihood after seeing $t$ samples -- or adding regularization, the MAP -- the maximum a posteriori.
As we will prove,\footnote{\fdk{I don't know any prior reference for this exact statement, but I'm not sure it's safe to claim this as a ``proof'' as if it was a new statement}} the MAP can be interpreted as online mirror descent with decreasing step-sizes.
Despite recent advances in this field, our convergence rates are still lacking; they are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.

@@

\paragraph{Conjugate Prior}
One conjugate prior \citep{agarwal2010geometric} for $p(X|\nat)$ is
\begin{align}
    p(\nat) &\propto \exp( - n_0 \bregman(\nat ; \nat_0) )
\end{align}
where $n_0$ and $\nat_0$ are (hyper)parameters of the prior, and $\bregman(\nat ; \nat_0)$ is introduced above.
Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\nat_0$.
The negative log-likelihood of the prior is
\begin{align*}
    -\log p(\nat) = n_0 (\logpart(\nat)  - \nat^\top \meanp_0 ) + \cst
\end{align*}
which is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \m_0, -n_0)$.
Then the joint log-likelihood of $\mD =(X_1,\dots,X_n)$ and $\nat$ is
\begin{multline}
    -\log p(\mD|\nat)p(\nat) 
    = (n_0+n) \logpart (\nat) \\
    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T_i \right) + \cst
    \label{eq:joint_likelihood}
\end{multline}
where $T_i=T(X_i)$.
This means the posterior given $\mD$ is part of the same family, with natural parameters $(n_0 \m_0 + \sum_i T_i , -(n_0 + n))$.

\paragraph{Maximum A Posteriori (MAP).}
Minimizing~\eqref{eq:joint_likelihood} over $\nat$ yields the MAP estimate
\begin{align}
    \hat \nat = \argmin_\nat -\log p(\mD|\nat) + n_0 \bregman(\nat ; \nat_0) \; .
\end{align}
If they can be satisfied, the MAP verifies the first oder optimality conditions
\begin{align}
    \nabla \logpart(\hat \nat_\text{MAP}) = \hat \meanp_\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
\end{align}
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
\begin{align}
	\hat \m_\text{MLE} = \frac{\sum_{i=1}^n T_i}{n}
\end{align}
These rules are known as moment matching.
The MLE and MAP estimates are statistics of the dataset $\mD$. Given a random dataset, we wish to bound their deviation from the optimum $\nat^*$ or $\meanp^*$.

@@

\paragraph{Remark.}
This could be seen as a concentration inequality, expressed with a Bregman divergence instead of a norm.
A key difference though, is that the random variable $T(X)$ is connected to the metric $\logpart$. 
Indeed expressions~\eqref{eq:bregmanMLE} or~\eqref{eq:bregmanMAP} can be infinite for another choice of random variable. 
For instance, if we plug in $\conj(\m)= -\log(\m)$, which defines a divergence on positive numbers, and $T(X) \sim \cN(0,1)$ which can be negative.

@@

\subsection{MAP as Stochastic Mirror Descent}
\TODO{s
}

@@

\subsection{Locally Quadratic Case}
All Bregman divergences are locally quadratic. 
Under some assumptions, we can quantify  when this quadratic behavior  kicks in.
For instance {\bf when $\conj$ is self-concordant} \citep[Ch.4.1]{nesterov2003introductory}. 
This is true when $T$ is 1-dimensional and $\logpart$ is self-concordant 
-- e.g. exponential distribution, 
Pareto distribution with known minimum value, 
Weibull distribution with known shape k, 
Laplace distribution
with known mean
-- or when $T$ lives in a compact \citep{bubeck2015entropic}. Then the bound
\begin{align}
	 \bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF^*}^2
\end{align}
holds whenever $\norm{\m^*-\m}_{\mF^*} < 0.21$. 
Similarly, \citet{kakade2010learning} assumes a bound on all higher order moments at $\nat^*$ to quantify when $\logpart$ starts behaving quadratically, but their rate does not directly apply to $\conj$ and $\m$.

\TODO{Reference \citet{ostrovskii2021finite}.}

@@

\section{"Burn In" Statistical Analysis}
Self-concordant line of work : 
\citet{bubeck2015entropic}
\citet{kakade2010learning},
\citet{anastasiou2017bounds},
\citet{marteauferey2019beyond},
\citet{ostrovskii2021finite}.

>>>>

\newcommand{\MAPm}{\hat \m_n}
\newcommand{\MAPt}{\hat \nat_n}
\DeclareMathSymbol{\shortminus}{\mathbin}{AMSa}{"39}

@@

There are algorithms for this setting, for example the MLE -- the (online) maximum likelihood after seeing $t$ samples -- or adding regularization, the MAP -- the maximum a posteriori.
As we show, the MAP can be interpreted as online mirror descent with decreasing step-sizes.
Despite recent advances in this field, our convergence rates are still lacking; they are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.

@@

\paragraph{Conjugate Prior}
One conjugate prior \citep{agarwal2010geometric} for $p(X|\nat)$ is
\begin{align}
    p(\nat) 
    &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \\
    &\propto \exp(n_0 \m_0^\top \nat - n_0 \logpart(\nat))
\end{align}
where $n_0$ and $\nat_0$ are (hyper)parameters of the prior.
This is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \m_0, -n_0)$.
Intuitively, $n_0$ is a number of fictive data points observed from a distribution with parameter $\nat_0$.

\paragraph{Maximum A Posteriori (MAP).}
Using Bayes rule, the posterior of $\nat$ given a dataset $\mD_n =(X_1,\dots,X_n)$ is
\begin{align}
	p(\nat | \mD_n)
	%\propto p(\mD|\nat)p(\nat) 
    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt))
    \label{eq:joint_likelihood}
\end{align}
reaching its maximum in $\MAPt$, the MAP estimate verifying the first order optimality conditions
\begin{align}
    \nabla \logpart(\MAPt) = \MAPm
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
\end{align}
where $T_i=T(X_i)$.
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
\begin{align}
	\hat \m_n^\text{MLE} = \frac{\sum_{i=1}^n T_i}{n}
\end{align}
This last equation is known as moment matching.
The MLE and MAP estimates are statistics of the dataset $\mD_n$. 
Given a random dataset, we wish to bound their deviation from the optimum $\nat^*$ or $\meanp^*$.

@@

\paragraph{Remark.}
This could be seen as a concentration inequality, expressed with a Bregman divergence instead of a norm.
However there is a connection between the random variable $T(X)$ and the metric $\logpart$. 
Indeed expressions~\eqref{eq:bregmanMLE} or~\eqref{eq:bregmanMAP} can be infinite for another choice of random variable. 
For instance, if we plug in $\conj(\m)= -\log(\m)$, which defines a divergence on positive numbers, and $T(X) \sim \cN(0,1)$ which can be negative.

@@

\subsection{MAP as Stochastic Mirror Descent}
The posterior~\eqref{eq:joint_likelihood} can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$, so that the MAP verifies
\begin{align*}
	\hat \nat_{n}
	= \argmin_\nat - \log p(X_{n} | \nat) + (n_0 + n -1) \bregman(\nat; \hat \nat_{n-1})
    %\label{eq:SBPP}
\end{align*}
which is the update of a Stochastic Bregman Proximal Point update with step-size $\inv{n_0 + n - 1}$. Expanding $p(X_{n} | \nat)$ and  playing with the Bregman we get
\begin{align}
		\hat \nat_{n}
	= \argmin_\nat g_{n}^\top \nat + (n_0 + n) \bregman(\nat; \hat \nat_{n-1})
    \label{eq:SMD}
\end{align}
where $g_{n} := \nabla\logpart(\hat \nat_{n-1}) - T(X_n) = \hat \m_{n-1} - T_n$ is the gradient of $-\log p(X_{n} | \nat)$ evaluated at $\hat \nat_{n-1}$. 
We recognize the update of Stochastic Mirror Descent -- a.k.a. Stochastic Bregman Gradient -- with step-size $\lr_n = \inv{n_0 + n}$.

@@

\subsection{Locally Quadratic Case}
All Bregman divergences are locally quadratic. 
Under some assumptions, we can quantify  when this quadratic behavior  kicks in.
For instance {\bf when $\conj$ is self-concordant} \citep[Ch.4.1]{nesterov2003introductory}. 
This is true when $T$ is 1-dimensional and $\logpart$ is self-concordant 
-- e.g. exponential distribution, 
Gaussian and Laplace distributions with known mean,
Pareto distribution with known minimum value, 
Weibull distribution with known shape k
-- or when $T$ lives in a compact \citep{bubeck2015entropic}. Then the bound
\begin{align}
	 \bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF^*}^2
\end{align}
holds whenever $\norm{\m^*-\m}_{\mF^*} < 0.21$. 
Similarly, \citet{kakade2010learning} assumes a bound on all higher order moments at $\nat^*$ to quantify when $\logpart$ starts behaving quadratically, but their rate does not directly apply to $\conj$ and $\m$.

\TODO{Reference \citet{ostrovskii2021finite}.} ( as well as 
\citet{anastasiou2017bounds},
\citet{marteauferey2019beyond})

@@
